---
title: "phyloseq-object-attempt"
output: html_document
---
#loading libraries

```{r}
library("readr")
library("readxl")
library("tidyr")
library("dplyr")
library("phyloseq")
library("stringr")
library("ggplot2")
library("microbiome")
library("vegan")
library("tibble")
```

#importing data

```{r}
ASV_raw <- read_tsv('/Users/chloecamacho/Desktop/thesis/Bioinformatics/raw data/library_g3_trim_prok_asvs.df.tsv')
TAXA <- read_tsv('/Users/chloecamacho/Desktop/thesis/Bioinformatics/raw data/library_g3_trim_prok_asvs.taxa.tsv')
meta <- read_excel("~/Desktop/thesis/Bioinformatics/meta?.xlsx")
contam_ASV <- read.delim("/Users/chloecamacho/Desktop/thesis/Bioinformatics/raw data/potential_contam_asvs.camacho.taxa.tsv", header = TRUE, sep = "\t")
```

##Cleaning data up

#getting rid of contaminant ASVs
```{r}

tax_cols <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")  

# unique contaminated lineages
bad_lineage <- contam_ASV %>% 
  select(all_of(tax_cols)) %>% 
  distinct()

# matching contaminated lineages in TAXA
matching_taxa <- TAXA %>%
  inner_join(bad_lineage, by = tax_cols, suffix = c("_main", "_bad")) %>%
  pull(asv_id)  # get ASV IDs to remove

```


#Creating matrices without contaminant ASVs, reordering ASV and TAXA, 

```{r}
ASV <- ASV_raw %>% 
  pivot_wider(
    id_cols    = asv_id,          # the rows = ASVs
    names_from = sample_name,     # the columns = samples
    values_from = counts,
    values_fill = 0
  ) 
#saving dfs as matrices
ASV_matrix <- ASV %>%
  filter(!asv_id %in% matching_taxa) %>% #removing bad ASVs
  column_to_rownames("asv_id") %>%   #making asv_id as the row name 
  mutate(across(everything(), as.numeric)) %>% #converting to numeric
  as.matrix()

TAXA_MAT <- TAXA %>%
  column_to_rownames("asv_id") %>%
  as.matrix()

#double checks
any(rownames(ASV_matrix) %in% matching_taxa) #double check that the matrix has no bad ASVs
class(ASV_matrix) #double check its a matrix
class(TAXA_MAT)

#cleaning up metadata
meta <- meta %>% 
  distinct(CamachoName, .keep_all = TRUE) %>% #removes duplicates
  rename(SamplingDate = `Sampling Date`) %>%
  mutate(CamachoName = str_replace(CamachoName, "PSV", "PSU")) #corrects PSV to PSU

#fixing rows
meta_df <- as.data.frame(meta) #saving as dataframe
rownames(meta_df) <- meta_df$CamachoName #making CamachoName column be the rows instead
meta_df$CamachoName <- NULL  # remove the ID column (now rownames)

#Find common samples & asvs and align ORDER
common_samples <- intersect(colnames(ASV_matrix), rownames(meta_df)) 
common_asvs <- intersect(rownames(ASV_matrix), rownames(TAXA_MAT))
length(common_asvs) 

# Subset both to matching samples & asvs, SAME ORDER
ASV_matrix <- ASV_matrix[, common_samples, drop = FALSE]
meta_df    <- meta_df[common_samples, , drop = FALSE]

ASV_matrix  <- ASV_matrix[common_asvs, , drop = FALSE]
TAXA_MAT    <- TAXA_MAT[common_asvs, , drop = FALSE]

# double checking everything is good 
identical(colnames(ASV_matrix), rownames(meta_df))  # Must be TRUE
identical(rownames(ASV_matrix), rownames(TAXA_MAT))  # TRUE

#from this point on im using ASV_matrix, TAXA_MAT, meta_df
```

#Creating phyloseq object

```{r}
OTU <- otu_table(ASV_matrix, taxa_are_rows = TRUE)
TAX = tax_table(TAXA_MAT)
META <- sample_data(meta_df)

physeq <- phyloseq(OTU,TAX, META)
physeq
```

#saving phyloseq object

```{r}
# Save your physeq object as an RDS file (compressed, fast to reload)
saveRDS(physeq, "/Users/chloecamacho/Desktop/thesis/Bioinformatics/physeq_filtered.rds")
```

#Plotting

```{r}

#################################### PCA PLOT ######################################################
site_colors <- c("Brewer's Bay"= "turquoise", "Saba"="lightpink", "Perseverance"="navy", "Krum Bay"="darkorange", "Rupert's Rock"="brown")

coral_colors <- c("Acropora palmata"="yellow", "Montastraea cavernosa"="darkorchid","Porites astreoides"="green")

sample_data(physeq)$Site <-factor(sample_data(physeq)$Site, levels = c("Brewer's Bay", "Saba", "Perseverance", "Krum Bay", "Rupert's Rock"))

physeq_sites <- subset_samples(physeq, Site %in% c("Brewer's Bay", "Saba", "Perseverance", "Krum Bay", "Rupert's Rock"))%>%
  filter_taxa(function(x) sum(x) > 0, TRUE)  # removes 35 taxa

# Transform the data for PCA
physeq_sites.beta <- microbiome::transform(physeq_sites, 'clr') # centered log ratio transformation
physeq_sites.beta_euc <- ordinate(physeq_sites.beta, "RDA", "euclidean") # Redundancy Analysis, Euclidean for distance matrix to calculate dissimilarity between samples

# plot PCA
beta_site <- plot_ordination(physeq_sites.beta, physeq_sites.beta_euc, 
                                   type="samples", color="Site", shape = "Species") +
  coord_fixed() +
  geom_point(size = 4) +
  labs(color = "Site", shape = "Species") +
  theme_minimal() +
  scale_color_manual(values = site_colors) 

beta_site
```

testing for 
```{r}
# PERMANOVA: Does Site structure communities?
clr_meta <- as(sample_data(physeq_sites.beta), "data.frame")
clr_otu <- otu_table(physeq_sites.beta) %>% as.matrix()

euc_diss <- vegdist(clr_otu, method = "euclidean")

set.seed(10)
adonis2(formula = euc_diss ~ Site, data = clr_meta, permutations = 999)
```

#one species per all sites

```{r}
physeq_past <-subset_samples(physeq,Species == "Porites astreoides")
physeq_past.beta <- microbiome::transform(physeq_past, 'clr') # centered log ratio transformation
physeq_past.beta_euc <- ordinate(physeq_past.beta, "RDA", "euclidean")

# plot PCA
beta_past <- plot_ordination(physeq_past.beta, physeq_past.beta_euc, 
                                   type="samples", color="Site", shape = "Species") +
  coord_fixed() +
  geom_point(size = 4) +
  labs(color = "Site", shape = "Species") +
  theme_minimal() +
  scale_color_manual(values = site_colors) 

beta_past
```
#permanova to see if site has effect on past
```{r}

clr_meta <- as(sample_data(physeq_past.beta), "data.frame")
clr_otu <- otu_table(physeq_past.beta) %>% as.matrix()

#get dissimilarity matrix 
euc_diss <- vegdist(clr_otu, method = "euclidean")

# permanova test to evaluate if treatment significantly structures microbial community beta diversity
set.seed(10)
adonis2(formula = euc_diss ~ Site, data = clr_meta, permutations = 999) 

# R2: 0.82774
# F: 12.814
# Pr(>F): 0.001
```

#stacked bar plot of blocks at each site on family level

```{r}
sample_variables(physeq)  # see available columns
unique(sample_data(physeq)$Site)  # see exact values

#filtering to only look at block samples 
physeq_filtered <- physeq %>%
  subset_samples(Site != "UVI Nursery") %>%           # No nursery
  subset_samples(SampleType == "Block")               # Blocks only

# Agglomerate to Family level
physeq_fam <- tax_glom(physeq_filtered, taxrank = "Family")

# 2. Transform to relative abundance
physeq_fam_rel <- transform_sample_counts(physeq_fam, function(x) x/sum(x))

# Plot (sites on X-axis, families stacked by relative abundance)
plot_bar(physeq_fam_rel, fill = "Family") + 
  labs(x = "Sample", y = "Relative Abundance", 
       title = "Families On Blocks Across Sites") +
 #scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  facet_wrap(~ Site, scales = "free_x", nrow = 1) +  # One row of site panels
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        legend.position = "none",
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(size = 10, face = "bold"))

```

#just PAST stacked bar plot

```{r}

# 1. filter to just look at past in the wild 
physeq_past <- physeq %>%
  subset_samples(SampleType == "Coral") %>%
  subset_samples(Species == "Porites astreoides") %>% 
  subset_samples(Site != "UVI Nursery")

nsamples(physeq_past)

# 2. Agglomerate to Family level  
past_family <- tax_glom(physeq_past, taxrank = "Family") %>% 
  

# 3. Transform to relative abundance
past_family_RA <- transform_sample_counts(past_family, function(x) x/sum(x))

# 4. Plot ALL families for this species across sites
plot_bar(past_family_RA, fill = "Family") + 
  labs(x = "Sample", y = "Relative Abundance", 
       title = "Families for Porites astreoides Across Sites") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  facet_wrap(~ Site, scales = "free_x", nrow = 1) +  # ← Site panels, no x-axis labels
  theme(axis.text.x = element_blank(),  # ← Hide sample names completely
        axis.ticks.x = element_blank(),
        legend.position = "right") +
  guides(fill = guide_legend(ncol = 1))

```

